On the instability and degeneracy of deep learning models

### 2019-06-18 writing/paper-note/docs/note.tex
revision of note

\newcommand{\REP}{\mathrm{LREP}}
\newcommand{\DN}{\Delta_N}

\newcommand{\ma}{\mathrm{max}_{\boldsymbol \theta_N}}
\newcommand{\mi}{\mathrm{min}_{\boldsymbol \theta_N}}

\newcommand{\maa}{\mathrm{max}_{-\boldsymbol \theta_N}}
\newcommand{\mii}{\mathrm{min}_{-\boldsymbol \theta_N}}

@@

\newcommand{\elt}{A_{N}(\boldsymbol \theta_N) }
\newcommand{\Gam}{B_{N}(\boldsymbol \theta_N) }

\newcommand{\Gamc}{C_{N}(\boldsymbol \theta_N) }
\newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }

@@

\begin{abstract}
{A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. This instability is a property of the probability \ak{model, prescribed by a probability distributional form evaluated at a given configuration of parameters}. For correlated data structures found in several application areas, there is increasing interest in \ak{identifying} such sensitivity in model probability structure. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length \(N\) has a finite number of possible values \ak{that can be taken at each point}. A sequence of probability models results, indexed by \(N\), \ak{and an associated parameter sequence,} that accommodates data of expanding dimension. Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than \(N\). In this case, a one component change in the data sequence can shift probability by orders of magnitude. Also, as instability becomes more extreme, the resulting probability models are shown to tend to degeneracy, placing all their probability on potentially small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.}
{Degeneracy, Instability, Classification, Deep Learning, Graphical Models}
\end{abstract}
% end imaiai stuff

@@

In the model framework, let \(\boldsymbol X_N = (X_1, \dots, X_N)\) denote a collection of discrete random variables with a finite sample space, \(\mathcal{X}^N\), represented as some \(N\)-fold Cartesian product. That is, \(\mathcal{X}\) with \(|\mathcal{X}| < \infty\) denotes the set of potential outcomes for each single variable \(X_i\), so that the product space \(\mathcal{X}^N\) corresponds to values for the variables \(\boldsymbol X_N=(X_1,\ldots,X_N)\). For each \(N\), let \(P_{\boldsymbol \theta_N}\) denote a probability model on \(\mathcal{X}^N\), under which \(P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0\) is the probability of the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\). In this, we assume that the model support of \(P_{\boldsymbol \theta_N}\) is the sample space \(\mathcal{X}^N\). This framework produces probability models \(P_{\boldsymbol \theta_N}\), indexed by a \ak{defining} sequence of parameters \(\boldsymbol \theta_N\), to describe data \(\boldsymbol X_N\) of any given sample size \(N \geq 1\). For simplicity, we will refer to this distributional class as \emph{Finite Outcome Everywhere Supported (FOES)} models in the following. The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where \(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary integer-valued function \(q(\cdot) \geq 1\).

Section \ref{examples} provides some examples of FOES models encountered in graph/network analysis and machine learning (i.e., deep learning models). These are used as references for later illustrations. Section \ref{instability-results} then establishes several formal results for FOES models with regard to instability. \citet{schweinberger2011instability} originally developed instability results specific to a certain class of discrete exponential models. For similar exponential models with random networks, \citet{handcock2003assessing} studied model degeneracy, where a probability model places near complete mass on modes and may thereby narrow the feasible model outcomes. As findings here and from \citet{schweinberger2011instability} suggest, model instability and degeneracy may also be related by viewing degeneracy as an extreme, or limiting form, of instability. Our main results establish a broad characterization of model instability, appropriate across the whole FOES model class, that incorporates results of \citet{schweinberger2011instability} as a special case. We prescribe a general and simple condition for identifying instability in a FOES model sequence, which quantifies whether certain maximal probabilities in a FOES model are too extreme relative to the sample size \(N\). When these conditions are met, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme peaks and troughs across nearly identical outcomes. As the measure of model instability increases, probabilities from an unstable FOES model additionally increase in volatility and provably slide into degeneracy. Section \ref{implications} then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and MCMC-based simulations. As one potential remedy, suggestions are given for constraining model parameterizations to avoid the most problematic regions of the parameter space. Proofs of the main results appear in Appendix \ref{appendix-instab}.

@@

For random variables \(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an exponential \ak{model family} for \(\boldsymbol X\) with probability mass function given by
\begin{equation}
\label{eq:expo}
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector \(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural parameter function \(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed positive integers \(k\) and \(L\) denoting their dimensions. Above, \(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of sufficient statistics, while
\[
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
\]
denotes the normalizing function with parameter space \(\Theta_N\). The natural parameter function \(\eta (\boldsymbol \theta)\) has a linear form (i.e., \(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given \(L \times k\) matrix \(\bm{A}\)) in many common model formulations, though may also be nonlinear (e.g., curved exponential families). In the linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics \(\boldsymbol g_N(\boldsymbol x)\).

Such discrete \ak{exponential models} are special cases of the FOES models, as seen by defining \(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0\), \(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a parameter sequence \(\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k\). For example, if observations \(\boldsymbol X = (X_1,\ldots,X_N)\) correspond to \(N\) independent and identically distributed Bernoulli random variables, each indicating a binary \(0\)-\(1\) outcome, the resulting probabilities have exponential form \eqref{eq:expo} given by

@@

with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds ratio'' parameter \(\boldsymbol \theta_N \equiv \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}\). More generally, supposing \(\boldsymbol X =(X_1,\ldots,X_N)\) represent \(N\) independent trials, each assuming an outcome \(\{1,\ldots,k\}\) among \(k\) possibilities (e.g., a die roll), a multinomial distribution is given by

@@

with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\) involving a count \(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome \(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the indicator function, and parameters \(\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k\) defining log-probability ratios \(\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]\). In addition to such standard models for discrete independent data, exponential models of FOES type commonly arise with dependent spatial data \citep{besag1974spatial} and network/relational data \citep{wasserman1994social, handcock2003assessing}. For a random graph or network with, say, \(n\) nodes, consider \(N={n \choose 2}\) random edges where the \(i\)th edge is associated with a pair of nodes \(s_i \equiv \{v_i,u_i\}\) and a binary variable \(X_i\in\{0,1\}\) indicating presence/absence of an edge among the node pair \(s_i\), \(i=1,\ldots,N\). Here the length \(N\) of the edge variable sequence \(\boldsymbol X = (X_1,\ldots,X_N)\) increases as a function of node number \(n\) and corresponding exponential models often incorporate graph topographical features derived from \(\boldsymbol X\). As an example, consider a graph model of exponential/FOES form prescribed by

@@

involving the numbers of edges, 2-stars and triangles among an outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\), \(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively, along with \(k=3\) real parameters \(\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})\). For this network model \eqref{eq:mod2} in particular, as well as for more general models of form \eqref{eq:expo}, \citet{schweinberger2011instability} considered instability in such exponential models with sequences of fixed parameters \(\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\), \(N \geq 1\), of fixed dimension \(k\).

For model sequences \(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)\) of the exponential type \eqref{eq:expo}, such as those in \eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the parameter \(\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k\) necessarily remains the same for all sample sizes \(N \geq 1\) as the form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo} and the number of sufficient statistics \(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\). Consequently, \(\boldsymbol \theta_N\) lies in a parameter space of fixed Euclidean dimension \(k\). However, this aspect need not be true for other types of FOES models considered in Sections \ref{rbm} - \ref{deep-learning}, where instead the numbers of parameters and sufficient statistics commonly increase with the sample size \(N\).

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, with binary variables being most common \citep[cf.~][]{smolensky1986information}. A RBM architecture has two layers, hidden (\(\mathcal{H}\)) and visible (\(\mathcal{V}\)), with conditional independence within each layer. Let \(\boldsymbol X = (X_1,\ldots,X_N)\) denote the \(N\) random variables for visibles with support \(\mathcal{X}^N\) and \(\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})\) denote the \(N_\mathcal{H}\) random variables for hiddens with support \(\mathcal{X}^{N_\mathcal{H}}\) where \(\mathcal{X} = \{-1,1\}\). For \ak{specified} parameters \(\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\), \(\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N\), and \(\boldsymbol \theta_N^{\mathcal{HV}}\) as a real matrix with dimension \(N_\mathcal{H} \times N\), the RBM model for \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint probability mass function
\label{eq:RBM1}
\tilde{P}_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
with normalizing function
\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x\right].
Let \(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}\), with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote \ak{a given} parameter vector for the RBM, as indexed by the number \(N\) of visible random variables (which may differ from the actual lengths of these parameter vectors). The probability mass function for the visible variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from marginalizing the joint specification to yield
\label{eq:RBM2}
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
Here the baseline model \eqref{eq:RBM1} for hidden/visible variables is a linear exponential one in sufficient statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from \eqref{eq:RBM1}, but the form differs from the previous exponential models in \eqref{eq:expo} in that the lengths of parameters \(\boldsymbol \theta_N\) and statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to incorporate more visible variables. That is, in contrast to \eqref{eq:expo}, the natural parameter function involved in the RBM model \eqref{eq:RBM1}, as the identity mapping of the parameters \(\boldsymbol \theta_N\in\mathbb{R}^{q(N)}\), naturally grows in dimension \(q(N)\to \infty\) to accommodate visible variables \(X_1, \dots, X_N\) of increasing sample size \(N\to \infty\). Additionally, one may further arbitrarily choose the number \(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RBM model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the \(N\) visible variables \(\boldsymbol X\), and the number \(N_\mathcal{H}\) of hiddens may also potentially increase with \(N\). Because \(|\mathcal{X}| = 2\) and \(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all \(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2} for visibles \(\boldsymbol\) corresponds to a FOES model, while the joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\) is also a FOES model. As this example also indicates, any model formed by marginalizing a base FOES model class, such as the RBM joint specification \eqref{eq:RBM1}, is again a FOES model.

\hypertarget{deep-learning}{%
\subsection{Deep Learning}\label{deep-learning}}

@@

As a basic characteristic, S-unstable FOES model sequences have extremely sensitive probability structures. One aspect is that small changes in data configuration can lead to very large changes in probability. Consider, for example, the quantity given by
\[
\DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
\]
which represents the biggest log-probability ratio for a one-component change \ak{(i.e., the data vector differs in exactly one component value)} in data outcomes in a FOES model with parameter \(\boldsymbol \theta_N\). \ak{This is the smallest difference possible between two data vectors, making it a good notion of a small distance in the input space.} We then have the following result prescribing the behavior of \(\DN(\boldsymbol \theta_N)\) for S-unstable FOES models.

>>>>

\newcommand{\REP}{\mathrm{LREP}}
\newcommand{\DN}{\Delta_N}
\newcommand{\thetaidx}{q(N)}
\newcommand{\thetaN}{\boldsymbol \theta_{\thetaidx}}

\newcommand{\ma}{\mathrm{max}_{\thetaN}}
\newcommand{\mi}{\mathrm{min}_{\thetaN}}

\newcommand{\maa}{\mathrm{max}_{-\thetaN}}
\newcommand{\mii}{\mathrm{min}_{-\thetaN}}

@@

\newcommand{\elt}{A_{N}(\thetaN) }
\newcommand{\Gam}{B_{N}(\thetaN) }

\newcommand{\Gamc}{C_{N}(\thetaN) }
\newcommand{\Gamt}{\Gamma_{N,2}(\thetaN) }

@@

\begin{abstract}
{A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. This instability is a property of the probability \ak{model, given by a distributional form and a given configuration of parameters}. For correlated data structures found in several application areas, there is increasing interest in \ak{identifying} such sensitivity in model probability structure. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length \(N\) has a finite number of possible values \ak{that can be taken at each point}. A sequence of probability models results, indexed by \(N\), \ak{and an associated parameter sequence,} that accommodates data of expanding dimension. Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than \(N\). In this case, a one component change in the data sequence can shift probability by orders of magnitude. Also, as instability becomes more extreme, the resulting probability models are shown to tend to degeneracy, placing all their probability on potentially small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.}
{Degeneracy, Instability, Classification, Deep Learning, Graphical Models}
\end{abstract}
% end imaiai stuff

@@

In the model framework, let \(\boldsymbol X_N = (X_1, \dots, X_N)\) denote a collection of discrete random variables with a finite sample space, \(\mathcal{X}^N\), represented as some \(N\)-fold Cartesian product. That is, \(\mathcal{X}\) with \(|\mathcal{X}| < \infty\) denotes the set of potential outcomes for each single variable \(X_i\), so that the product space \(\mathcal{X}^N\) corresponds to values for the variables \(\boldsymbol X_N=(X_1,\ldots,X_N)\). For each \(N\), let \(P_{\ak{\thetaN}}\) denote a probability model on \(\mathcal{X}^N\), under which \(P_{\ak{\thetaN}}(x_1, \dots, x_N) > 0\) is the probability of the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\). In this, we assume that the model support of \(P_{\ak{\thetaN}}\) is the sample space \(\mathcal{X}^N\). This framework produces probability models \(P_{\ak{\thetaN}}\), indexed by a \ak{defining} sequence of parameters \(\ak{\thetaN}\), to describe data \(\boldsymbol X_N\) of any given sample size \(N \geq 1\). For simplicity, we will refer to this distributional class as \emph{Finite Outcome Everywhere Supported (FOES)} models in the following. The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where \(\ak{\thetaN} \in \mathbb{R}^{q(N)}\) for some arbitrary integer-valued function \(q(\cdot) \geq 1\).

Section \ref{examples} provides some examples of FOES models encountered in graph/network analysis and machine learning (i.e., deep learning models). These are used as references for later illustrations. Section \ref{instability-results} then establishes several formal results for FOES models with regard to instability. \citet{schweinberger2011instability} originally developed instability results specific to a certain class of discrete exponential models. For similar exponential models with random networks, \citet{handcock2003assessing} studied model degeneracy, where a probability model places near complete mass on modes and may thereby narrow the feasible model outcomes. As findings here and from \citet{schweinberger2011instability} suggest, model instability and degeneracy may also be related by viewing degeneracy as an extreme, or limiting form, of instability. Our main results establish a broad characterization of model instability, appropriate across the whole FOES model class, that incorporates results of \citet{schweinberger2011instability} as a special case. We prescribe a general and simple condition for identifying instability in a FOES model sequence, which quantifies whether certain maximal probabilities in a FOES model are too extreme relative to the sample size \(N\). When these conditions are met, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme peaks and troughs across nearly identical outcomes. As the measure of model instability increases, probabilities from an unstable FOES model additionally increase in volatility and provably slide into degeneracy. Section \ref{implications} then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and MCMC-based simulations. As one potential remedy, suggestions are given for constraining model parameterizations to avoid the most problematic regions of the parameter space \ak{(see Section \ref{bayes} for further discussion)}. Proofs of the main results appear in Appendix \ref{appendix-instab}.

@@

For random variables \(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an exponential \ak{model family} for \(\boldsymbol X\) with probability mass function given by
\begin{equation}
\label{eq:expo}
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector \(\boldsymbol \theta \in \Theta_{\ak{\thetaidx}} \subset \mathbb{R}^{k}\) and natural parameter function \(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed positive integers \(k\) and \(L\) denoting their dimensions. Above, \(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of sufficient statistics, while
\[
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_\ak{\thetaidx}\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
\]
denotes the normalizing function with parameter space \(\Theta_{\ak{\thetaidx}}\). The natural parameter function \(\eta (\boldsymbol \theta)\) has a linear form (i.e., \(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given \(L \times k\) matrix \(\bm{A}\)) in many common model formulations, though may also be nonlinear (e.g., curved exponential families). In the linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics \(\boldsymbol g_N(\boldsymbol x)\).

Such discrete \ak{exponential models} are special cases of the FOES models, as seen by defining \(P_{\ak{\thetaN}}(\boldsymbol x)\equiv p_{N, \ak{\thetaN}}(\boldsymbol x)> 0\), \(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a parameter sequence \(\ak{\thetaN} \in \Theta_{\ak{\thetaidx}} \subset \mathbb{R}^k\). For example, if observations \(\boldsymbol X = (X_1,\ldots,X_N)\) correspond to \(N\) independent and identically distributed Bernoulli random variables, each indicating a binary \(0\)-\(1\) outcome, the resulting probabilities have exponential form \eqref{eq:expo} given by

@@

with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds ratio'' parameter \(\ak{\thetaN} \equiv \log[ P_{\ak{\thetaN}}(X_i=1)/P_{\ak{\thetaN}}(X_i=0) ] \in \mathbb{R}\). More generally, supposing \(\boldsymbol X =(X_1,\ldots,X_N)\) represent \(N\) independent trials, each assuming an outcome \(\{1,\ldots,k\}\) among \(k\) possibilities (e.g., a die roll), a multinomial distribution is given by

@@

with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\) involving a count \(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome \(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the indicator function, and parameters \(\ak{\thetaN}=(\theta_{1,\ak{\thetaidx}},\ldots,\theta_{k,\ak{\thetaidx}})\in\mathbb{R}^k\) defining log-probability ratios \(\theta_{i,\ak{\thetaidx}}-\theta_{j,\ak{\thetaidx}} =\log [P_{\ak{\thetaN}}(X_1=i)/P_{\ak{\thetaN}}(X_1=j)]\). In addition to such standard models for discrete independent data, exponential models of FOES type commonly arise with dependent spatial data \citep{besag1974spatial} and network/relational data \citep{wasserman1994social, handcock2003assessing}. For a random graph or network with, say, \(n\) nodes \ak{$\{v_i\}_{i = 1}^n$}, consider \(N={n \choose 2}\) random edges where the \(i\)th edge is associated with a pair of nodes \(s_i \equiv \{v_{i_1},v_{i_2}\}\) and a binary variable \(X_i\in\{0,1\}\) indicating presence/absence of an edge among the node pair \(s_i\), \(i=1,\ldots,N\). Here the length \(N\) of the edge variable sequence \(\boldsymbol X = (X_1,\ldots,X_N)\) increases as a function of node number \(n\) and corresponding exponential models often incorporate graph topographical features derived from \(\boldsymbol X\). As an example, consider a graph model of exponential/FOES form prescribed by

@@

involving the numbers of edges, 2-stars and triangles among an outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\), \(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively, along with \(k=3\) real parameters \(\ak{\thetaN} \equiv (\theta_{1,\ak{\thetaidx}},\theta_{2,\ak{\thetaidx}},\theta_{3,\ak{\thetaidx}})\). For this network model \eqref{eq:mod2} in particular, as well as for more general models of form \eqref{eq:expo}, \citet{schweinberger2011instability} considered instability in such exponential models with sequences of fixed parameters \(\ak{\thetaN} = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\), \(N \geq 1\), of fixed dimension \(k\).

For model sequences \(P_{\ak{\thetaN}}(\boldsymbol x)\equiv p_{N,\ak{\thetaN}}(\boldsymbol x)\) of the exponential type \eqref{eq:expo}, such as those in \eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the parameter \(\ak{\thetaN}\in\Theta \subset \mathbb{R}^k\) necessarily remains the same for all sample sizes \(N \geq 1\) as the form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo} and the number of sufficient statistics \(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\). Consequently, \(\ak{\thetaN}\) lies in a parameter space of fixed Euclidean dimension \(k\). However, this aspect need not be true for other types of FOES models considered in Sections \ref{rbm} - \ref{deep-learning}, where instead the numbers of parameters and sufficient statistics commonly increase with the sample size \(N\).

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, with binary variables being most common \citep[cf.~][]{smolensky1986information}. A RBM architecture has two layers, hidden (\(\mathcal{H}\)) and visible (\(\mathcal{V}\)), with conditional independence within each layer. Let \(\boldsymbol X = (X_1,\ldots,X_N)\) denote the \(N\) random variables for visibles with support \(\mathcal{X}^N\) and \(\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})\) denote the \(N_\mathcal{H}\) random variables for hiddens with support \(\mathcal{X}^{N_\mathcal{H}}\) where \(\mathcal{X} = \{-1,1\}\). For \ak{specified} parameters \(\ak{\thetaN}^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\), \(\ak{\thetaN}^{\mathcal{V}}\in \mathbb{R}^N\), and \(\ak{\thetaN}^{\mathcal{HV}}\) as a real matrix with dimension \(N_\mathcal{H} \times N\), the RBM model for \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint probability mass function
\begin{equation}
\label{eq:RBM1}
\tilde{P}_{\ak{\thetaN}} (\tilde{\boldsymbol x}) = \exp\left[ (\ak{\thetaN}^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\ak{\thetaN}^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \ak{\thetaN}^{\mathcal{HV}} \boldsymbol x - \psi(\ak{\thetaN})\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
\end{equation}
with \ak{normalizing function $\psi(\ak{\thetaN})$}.
Let \(\ak{\thetaN} = (\ak{\thetaN}^{\mathcal{H}}, \ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{HV}} ) \in \Theta_\ak{\thetaidx} \equiv \mathbb{R}^{q(N)}\), with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote \ak{a given} parameter vector for the RBM, as indexed by the number \(N\) of visible random variables (which may differ from the actual lengths of these parameter vectors). The probability mass function for the visible variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from marginalizing the joint specification to yield
\begin{equation}
\label{eq:RBM2}
P_{\ak{\thetaN}} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
\end{equation}
Here the baseline model \eqref{eq:RBM1} for hidden/visible variables is a linear exponential one in sufficient statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from \eqref{eq:RBM1}, but the form differs from the previous exponential models in \eqref{eq:expo} in that the lengths of parameters \(\ak{\thetaN}\) and statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to incorporate more visible variables. That is, in contrast to \eqref{eq:expo}, the natural parameter function involved in the RBM model \eqref{eq:RBM1}, as the identity mapping of the parameters \(\ak{\thetaN}\in\mathbb{R}^{q(N)}\), naturally grows in dimension \(q(N)\to \infty\) to accommodate visible variables \(X_1, \dots, X_N\) of increasing data size \(N\to \infty\). Additionally, one may further arbitrarily choose the number \(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RBM model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the \(N\) visible variables \(\boldsymbol X\), and the number \(N_\mathcal{H}\) of hiddens may also potentially increase with \(N\). Because \(|\mathcal{X}| = 2\) and \(P_{\ak{\thetaN}}(\boldsymbol x) > 0\) for all \(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2} for visibles \(\boldsymbol\) corresponds to a FOES model, while the joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\) is also a FOES model. As this example also indicates, any model formed by marginalizing a base FOES model class, such as the RBM joint specification \eqref{eq:RBM1}, is again a FOES model.

\ak{\subsection{Binary Deep Learning Models} \label{deep-learning}}

@@

As a basic characteristic, S-unstable FOES model sequences have extremely sensitive probability structures. One aspect is that small changes in data configuration can lead to very large changes in probability. Consider, for example, the quantity given by
\[
\DN(\ak{\thetaN}) \equiv \max \left\{\log \frac{P_{\ak{\thetaN}}(\boldsymbol x)}{P_{\ak{\thetaN}}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
\]
which represents the biggest log-probability ratio for a one-component change \ak{(i.e., the data vector differs in exactly one component value)} in data outcomes in a FOES model with parameter \(\ak{\thetaN}\). \ak{This is the smallest difference possible between two data vectors, making it a good notion of a small distance in the input space and has been similarly used by \citet[][, Sec 2.2]{schweinberger2011instability}.} We then have the following result prescribing the behavior of \(\DN(\ak{\thetaN})\) for S-unstable FOES models.
